---
title: "Introduction to cori.data.qcew"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to cori.data.qcew}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 9,
  fig.height = 6,
  out.width  = "100%"
)
```

`cori.data.qcew` provides access to Bureau of Labor Statistics Quarterly Census
of Employment and Wages (QCEW) data processed and hosted by CORI. The package
covers county, state, and national employment and wages from **1990 to present**.

## Reading data from S3

The primary function is `read_qcew_from_s3()`. By default it returns the latest
vintage, all geographies, all years, and all variables.

```{r read-basic}
library(cori.data.qcew)

df <- read_qcew_from_s3()
```

In practice you will almost always filter by variable, years, and geography:

```{r read-filtered}
# Employment for all West Virginia counties, 2010-2024
df <- read_qcew_from_s3(
  variables = "annual_avg_emplvl",
  geoids    = "54",          # 2-digit state FIPS returns all WV rows
  years     = 2010:2024
)
```

### Geoid conventions

| `geoid` | Meaning |
|---|---|
| `"00"` | National total |
| `"54"` | State total (West Virginia) |
| `"54039"` | County (Kanawha County, WV) |

```{r read-mixed}
df <- read_qcew_from_s3(
  geoids    = c("00", "54039"),
  variables = "annual_avg_emplvl"
)
```

## Data structure

Every row is one geoid / year / variable observation:

```{r structure}
dplyr::glimpse(df)
#> Rows: ...
#> $ geoid    <chr>  "54039", "54039", ...
#> $ year     <int>  2023, 2023, ...
#> $ variable <chr>  "annual_avg_emplvl", "annual_avg_pay", ...
#> $ value    <dbl>  97110, 52840, ...
#> $ agg_var  <dbl>  NA, 97110, ...
```

`agg_var` is the employment weight used for computing employment-weighted
averages of wage variables upstream. It is `NA` for employment counts.

## Codebook

`get_qcew_codebook()` documents every variable:

```{r codebook}
cb <- get_qcew_codebook()
cb
```

| Variable | Label | Unit |
|---|---|---|
| `annual_avg_emplvl` | Average annual employment | workers |
| `tradable_goods_emp_share` | Tradable goods employment share | proportion (0–1) |
| `tradable_services_emp_share` | Tradable services employment share | proportion (0–1) |
| `local_services_emp_share` | Local services employment share | proportion (0–1) |
| `annual_avg_pay` | Average annual pay | dollars/worker (nominal) |
| `tradable_goods_avg_annual_pay` | Tradable goods avg. annual pay | dollars/worker (nominal) |
| `tradable_services_avg_annual_pay` | Tradable services avg. annual pay | dollars/worker (nominal) |
| `local_services_avg_annual_pay` | Local services avg. annual pay | dollars/worker (nominal) |
| `hhi_emp` | Employment HHI | index (0–10,000) |

CORI super-sector definitions:

| Super-sector | BLS NAICS industries included |
|---|---|
| Tradable goods | Natural resources and mining, Construction, Manufacturing, Trade/transportation/utilities |
| Tradable services | Information, Financial activities, Professional and business services |
| Local services | Education and health services, Leisure and hospitality, Other services, Public administration |

## Vintage management

Each vintage is a complete snapshot of all years. The `_LATEST` pointer on S3
always points to the most recently published vintage.

```{r vintage}
# Check the current vintage
latest_qcew_vintage()
#> [1] "vintage_2024"

# Read a specific vintage explicitly
df_2024 <- read_qcew_from_s3(vintage = "2024")
```

## A simple example

```{r example}
library(cori.data.qcew)
library(cori.charts)
library(dplyr)
library(ggplot2)

load_fonts()

kanawha <- read_qcew_from_s3(
  geoids    = "54039",
  variables = "annual_avg_emplvl",
  years     = 1990:2024
)

ggplot(kanawha, aes(x = year, y = value)) +
  geom_line(linewidth = 1.2, color = cori_colors[["Emerald"]]) +
  scale_y_continuous(
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_x_continuous(
    expand   = expansion(mult = c(0, 0)),
    n.breaks = 6
  ) +
  theme_cori() +
  labs(
    title   = "Average Annual Employment — Kanawha County, WV",
    x       = NULL,
    y       = NULL,
    caption = "Source: BLS Quarterly Census of Employment and Wages via cori.data.qcew"
  )
```

## Further reading

- `vignette("rural-employment-trends")` — comparing rural and nonrural
  employment since the Great Recession using `ruraldefinitions`
