---
title: "Rural vs. Nonrural Employment Trends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rural vs. Nonrural Employment Trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.width  = 9,
  fig.height = 6,
  out.width  = "100%"
)
```

This vignette shows a common analysis pattern: comparing employment trends
between rural and nonrural counties since the Great Recession using
`read_qcew_from_s3()`, the `ruraldefinitions` package, and `cori.charts` for
visualization.

```{r load}
library(cori.data.qcew)
library(cori.charts)
library(ruraldefinitions)
library(ggrepel)
library(dplyr)
library(ggplot2)

load_fonts()
```

## Rural vs. nonrural employment since the Great Recession

Pull county-level annual employment for all years since 2007:

```{r read}
emp_raw <- read_qcew_from_s3(
  variables = "annual_avg_emplvl",
  years     = 2007:2024
)
```

Keep only 5-digit county geoids and attach the rural/nonrural flag from
`ruraldefinitions::cbsa_2023`:

```{r join}
emp_co <- emp_raw |>
  filter(nchar(geoid) == 5) |>
  left_join(
    select(cbsa_2023, geoid, is_rural),
    by = "geoid"
  )
```

Index employment to 2007 levels within each group to show relative change:

```{r index}
chrt_dta <- emp_co |>
  filter(!is.na(is_rural)) |>
  group_by(is_rural, year) |>
  reframe(tot_emp = sum(value)) |>
  group_by(is_rural) |>
  mutate(
    index_value = tot_emp / tot_emp[year == 2007],
    year        = as.numeric(year)
  )
```

Quick stats for the most recent year-over-year change:

```{r stats}
latest_yr <- max(chrt_dta$year)

chrt_dta |>
  filter(year >= (latest_yr - 1)) |>
  arrange(is_rural, year) |>
  group_by(is_rural) |>
  mutate(pct_change = (tot_emp - lag(tot_emp)) / lag(tot_emp)) |>
  filter(year == latest_yr) |>
  select(is_rural, pct_change)
```

```{r line-chart}
label_df <- chrt_dta |> filter(year == max(year))

chrt_dta |>
  ggplot(aes(x = year, y = index_value, color = is_rural, group = is_rural)) +
  annotate("rect", xmin = 2007, xmax = 2009, ymin = .9, ymax = 1.2,
           fill = "grey90", alpha = 0.5) +
  geom_line(linewidth = 1.5) +
  geom_text_repel(
    data          = label_df,
    aes(label     = is_rural),
    nudge_x       = 2,
    direction     = "y",
    hjust         = 0,
    segment.color = NA,
    size          = 5
  ) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1),
    limits = c(.9, 1.2),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_x_continuous(
    expand   = expansion(mult = c(0, 0)),
    n.breaks = 5,
    labels   = scales::label_number(accuracy = 1, big.mark = "")
  ) +
  scale_color_manual(
    values = c(
      "Rural"    = cori_palettes$ctg2ruralnonrural[[1]],
      "Nonrural" = cori_palettes$ctg2ruralnonrural[[2]]
    )
  ) +
  theme_cori() +
  theme(
    legend.position = "none",
    axis.ticks.x    = element_line(color = "#d0d2ce", linewidth = .25),
    plot.title      = element_text(color = cori_colors[["Dark Green"]])
  ) +
  labs(
    title    = "A fragile recovery: Rural employment set back twice since 2007",
    subtitle = "Relative change in employment since the Great Recession",
    x        = NULL,
    y        = NULL,
    caption  = paste0(
      "Source: CORI analysis of BLS Quarterly Census of Employment and Wages.\n",
      "Note: Indexed to 2007 levels. Rural = nonmetro counties per CBSA 2023 definition."
    )
  )
```

## Regional breakdown

To see how rural employment trends differ by Census region, attach a region
crosswalk and repeat the analysis:

```{r regional}
region_xwalk <- tigris::states() |>
  sf::st_drop_geometry() |>
  select(GEOID, REGION) |>
  mutate(
    region = case_when(
      REGION == "1" ~ "Northeast",
      REGION == "2" ~ "Midwest",
      REGION == "3" ~ "South",
      REGION == "4" ~ "West"
    )
  )

regional_rural <- emp_co |>
  filter(!is.na(is_rural), is_rural == "Rural") |>
  mutate(st_fips = substr(geoid, 1, 2)) |>
  left_join(region_xwalk, by = c("st_fips" = "GEOID")) |>
  filter(!is.na(region)) |>
  group_by(region, year) |>
  reframe(tot_emp = sum(value)) |>
  group_by(region) |>
  mutate(
    index_value = tot_emp / tot_emp[year == 2007],
    year        = as.numeric(year)
  )
```

```{r regional-chart}
region_colors <- c(
  "West"      = cori_palettes$ctg4mid[[1]],
  "South"     = cori_palettes$ctg4mid[[2]],
  "Midwest"   = cori_palettes$ctg4mid[[3]],
  "Northeast" = cori_palettes$ctg4mid[[4]]
)

label_df <- regional_rural |> filter(year == max(year))

regional_rural |>
  ggplot(aes(x = year, y = index_value, color = region)) +
  annotate("rect", xmin = 2007, xmax = 2009, ymin = -Inf, ymax = Inf,
           fill = "grey90", alpha = 0.5) +
  geom_hline(yintercept = 1, color = "grey70", linetype = "dashed") +
  geom_line(linewidth = 1.5) +
  geom_text_repel(
    data          = label_df,
    aes(label     = region),
    nudge_x       = 0.5,
    direction     = "y",
    hjust         = 0,
    segment.color = NA,
    size          = 4.5,
    show.legend   = FALSE
  ) +
  scale_color_manual(values = region_colors) +
  scale_x_continuous(
    limits = c(2007, max(regional_rural$year) + 3),
    expand = expansion(mult = c(0.02, 0)),
    labels = scales::label_number(accuracy = 1, big.mark = "")
  ) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1),
    expand = expansion(mult = c(0, .1))
  ) +
  theme_cori() +
  theme(
    legend.position = "none",
    axis.ticks.x    = element_line(color = "#d0d2ce", linewidth = .25),
    plot.title      = element_text(color = cori_colors[["Dark Green"]])
  ) +
  labs(
    title    = "Rural employment by Census region: West leads recovery",
    subtitle = "Relative change in rural employment since the Great Recession",
    x        = NULL,
    y        = NULL,
    caption  = paste0(
      "Source: CORI analysis of BLS Quarterly Census of Employment and Wages.\n",
      "Note: Indexed to 2007 levels. Rural = nonmetro counties per CBSA 2023 definition."
    )
  )
```

## Industry concentration: HHI swarm plot

The Herfindahl-Hirschman Index (`hhi_emp`) measures how concentrated a county's
employment is across industries. Higher values mean fewer industries dominate
local employment — a sign of economic fragility. Comparing rural and nonrural
counties in the Midwest illustrates how this plays out within a single region.

```{r hhi-read}
library(ggbeeswarm)

midwest_fips <- c("17", "18", "19", "20", "26", "27", "29", "31", "38", "39", "46", "55")

qcew <- read_qcew_from_s3(
  variables = c("hhi_emp", "annual_avg_emplvl"),
  years     = 2024
) |>
  filter(
    nchar(geoid) == 5,
    substr(geoid, 1, 2) %in% midwest_fips
  )

hhi <- qcew |>
  filter(variable == "hhi_emp") |>
  select(geoid, hhi = value)

emp <- qcew |>
  filter(variable == "annual_avg_emplvl") |>
  select(geoid, employment = value)

midwest_co <- hhi |>
  left_join(emp, by = "geoid") |>
  left_join(select(cbsa_2023, geoid, is_rural), by = "geoid") |>
  filter(!is.na(is_rural), !is.na(hhi))
```

```{r hhi-swarm}
midwest_co |>
  ggplot(aes(
    x     = hhi,
    y     = is_rural,
    size  = employment,
    color = is_rural
  )) +
  geom_beeswarm(
    cex      = 0.6,
    groupOnX = FALSE,
    alpha    = 0.4
  ) +
  scale_x_continuous(
    labels = scales::label_comma(),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_color_manual(
    values = c(
      "Rural"    = cori_palettes$ctg2ruralnonrural[[1]],
      "Nonrural" = cori_palettes$ctg2ruralnonrural[[2]]
    ),
    guide = "none"
  ) +
  scale_size(
    range  = c(0.5, 6),
    labels = scales::label_comma(scale_cut = scales::cut_short_scale()),
    guide  = guide_legend(
      title          = "Employment",
      title.position = "left",
      title.theme    = element_text(margin = margin(r = 8), size = 11),
      label.theme    = element_text(margin = margin(r = 5),  size = 11)
    )
  ) +
  theme_cori() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.ticks.x       = element_line(color = "#d0d2ce", linewidth = .25),
    plot.title         = element_text(color = cori_colors[["Dark Green"]])
  ) +
  labs(
    title    = "Rural Midwest counties have more concentrated economies",
    subtitle = "Employment HHI by rural status, Midwest counties (2024)",
    x        = "Employment HHI (higher = more concentrated)",
    y        = NULL,
    caption  = paste0(
      "Source: CORI analysis of BLS Quarterly Census of Employment and Wages.\n",
      "Note: Each dot is one county, sized by total employment. ",
      "Rural = nonmetro per CBSA 2023 definition."
    )
  )
```

## Notes

- The grey band marks the Great Recession (2007–2009).
- Rural counties are defined as nonmetro counties using the CBSA 2023
  definition from `ruraldefinitions::cbsa_2023`.
- Employment totals reflect covered employment (private + government) from the
  BLS QCEW all-industry total rows.
- HHI is computed across BLS NAICS super-sectors. Values above 2,500 are
  generally considered highly concentrated.
